---
title: "Reproducible Research Peer Assignement 1"
output: html_document
---

This is an assignment on answering a series of questions to practice loading data, cleaning it and interpreting data in a way that is clearly marked and presented on a html format. The data is collected from a [Fitbit](http://en.wikipedia.org/wiki/Fitbit).
## Data
The data for this assignment was downloaded from the course web
site:

Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K]

The variables included in this dataset are:

1. steps: Number of steps taking in a 5-minute interval (missing
    values are coded as `NA`)

2. date: The date on which the measurement was taken in YYYY-MM-DD format

3. interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.


## Part 1: Loading and preprocessing the data

```{r}
data <- read.csv("~/mm/activity.csv",  colClasses = c("numeric", "character", 
                                                          "numeric"))

```

## Part 2: What is mean total number of steps taken per day?
- Calculate the aggregate steps for all intervals sorted by day
- Plot Total Steps by Day
- Look for the interval with most steps.

```{r}
StepsTotal <- aggregate(steps ~ interval, data, mean)
dailysteps <- aggregate(steps~date,data,sum)

hist(dailysteps$steps, main = "Total Steps By Day", xlab = "Number of Steps", col = "yellow")

max_interval <- StepsTotal[which.max(StepsTotal$steps),1]
mean(dailysteps$steps)
median(dailysteps$steps)
```

##Part 3: What is the average daily activity pattern?

###Use time series plot to check 5 minute intervals
The 5-minute interval, on average across all the days in the data set, containing the maximum number of steps is `r max_interval`.

```{r}
time_series <- tapply(activity$steps, activity$interval, mean, na.rm = TRUE)

##And plot: 
plot(row.names(time_series), time_series, type = "l", xlab = "5-min interval", 
     ylab = "Average across all Days", main = "Average number of steps taken", 
     col = "red")
## to find max number of steps on average in a 5 min interval
max_interval <- which.max(time_series)
names(max_interval)
```

Therefore, the max number of steps taken on a 5 minute average in a day is 835.

##Part 4: Imputing missing values

###Find the number of missing values:

```{r}
empty <-sum(!complete.cases(data))
newdata <- transform(data, steps = ifelse(is.na(data$steps), StepsTotal$steps[match(data$interval, StepsTotal$interval)], data$steps))
newdata[as.character(newdata$date) == "2012-10-01", 1] <- 0
```
We altered the original data by replacing NA values as 0.

So now with the filled in data, we can recount the information given.

## To make it more comparable, here is a display of the changes that occurred due to the removal of "NA" figures, as shown in the blue parts in the histogram.
```{r}

dailysteps.2 <- aggregate(steps~date,newdata,sum)
hist(dailysteps.2$steps, main = "Total Steps by Day", xlab = "Number of Steps", col="blue")

StepsTotal2 <- aggregate(steps ~ date, data = new_activity, sum, na.rm = TRUE)
hist(StepsTotal2$steps, main = "Total steps by day", xlab = "day", col = "blue")


hist(dailysteps$steps, main = paste("Total Steps Each Day"), col="yellow", xlab="Number of Steps", add=T)
legend("topright", c("Original","Edited"), col=c("yellow", "blue"), lwd=10)

##New Mean and median

mean(dailysteps.2$steps)

median(dailysteps.2$steps)

```

To show the change in mean and median

```{r}
change1<- mean(dailysteps.2$steps) - mean(dailysteps$steps)
print(change1)
change2<- median(dailysteps.2$steps) - median(dailysteps$steps)
print(change2)

change3 <- sum(dailysteps.2$steps) - sum(dailysteps$steps)
print(change3)
```

##Changes after edit
` Mean changed by -176, 
` Median increased by 1.18. 
` There has also been 75363 steps added when the data is changed.


## Part 5: Are there differences in activity patterns between weekdays and weekends?
###Weekday versus weekend walking: differentiate between the two first, by sorting the data and adding a new column of either Weekday or Weekend
```{r}
weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
newdata$dow = as.factor(ifelse(is.element(weekdays(as.Date(newdata$date)),weekdays), "Weekday", "Weekend"))

stepsint<- aggregate(steps ~ interval + dow, newdata, mean)

```
##Create the plot:
```{r}
library(lattice)

xyplot(stepsint$steps ~ stepsint$interval|stepsint$dow, layout= c(1,2), xlab="Interval", ylab="Steps", main="Average Steps per day by interval", type = "l", col="pink")

```

